% =====================================================================
% Only Ancestors Matter: Ragged Flash Attention for
% Tree-Structured Speculative Decoding
% Target: NeurIPS 2026
% =====================================================================
\documentclass{article}

% ── NeurIPS 2026 style ──────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{microtype}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\usepackage{orcidlink}
\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=green!50!black,
  urlcolor=blue!70!black,
}

% ── Macros ──────────────────────────────────────────────────────────
\newcommand{\method}{TreeFlash}
\newcommand{\methodshort}{TreeFlash}
\newcommand{\bigO}[1]{\mathcal{O}(#1)}
\newcommand{\numtokens}{N}
\newcommand{\depth}{d}
\newcommand{\branching}{b}
\newcommand{\headdim}{D}
\newcommand{\nheads}{H}
\newcommand{\batchsize}{B}
\newcommand{\prefixlen}{L}
\DeclareMathOperator*{\softmax}{softmax}
\DeclareMathOperator{\parent}{parent}

\title{%
  \textbf{Draft Collapse: Tree-Sparse Attention for Scalable Speculative Decoding}%
}

\author{%
  Anonymous Author(s) \\
  Affiliation withheld for double-blind review
}

\date{}

\begin{document}
\maketitle

% =====================================================================
% ABSTRACT
% =====================================================================
\begin{abstract}
Tree-structured speculative decoding---used by
EAGLE-2/3~\cite{li2024eagle2,li2025eagle3},
Medusa~\cite{cai2024medusa}, and Sequoia~\cite{chen2024sequoia}---verifies
hundreds to thousands of draft tokens per target step.
During verification each draft token attends only to its
$\depth{+}1$ ancestors along the path from the tree root, yet production
attention kernels (FlashAttention-2~\cite{dao2023flashattention2},
FlashInfer~\cite{ye2025flashinfer}, DeFT~\cite{yao2024deft})
tile over all $\numtokens$ KV positions and burn $\bigO{\numtokens^2}$
work masking the rest to $-\infty$.

We present \textbf{\method{}}, a Triton ragged-attention kernel that reduces
per-head verification cost from $\bigO{\numtokens^2}$ to
$\bigO{\numtokens \cdot \depth}$ — \emph{linear} in tree size.
The kernel inverts the FlashAttention-2 inner loop: rather than streaming
KV blocks and masking, it walks each query's ancestor chain in BFS order
($\parent(k){=}\lfloor(k{-}1)/\branching\rfloor$), issuing $\depth{+}1$
scattered K/V loads per query row that feed a single-element online-softmax
update. We provide three variants — a closed-form b-ary kernel, an
explicit-parent kernel for EAGLE-3's dynamic/pruned trees, and a paged-KV
kernel compatible with PagedAttention~\cite{kwon2023efficient} serving — all
with an LSE-returning entry point for prefix-cache merging. Outputs match a
dense FA-2 reference to fp16/bf16 tolerance across the full test matrix.

Across kernel and end-to-end benchmarks on RTX 4000 Ada:
\textbf{(1)}~At $\prefixlen{=}0$, $\batchsize{=}1$, $\branching{=}32$,
$\depth{=}32$ (tree size ${\approx}880$), \method{} is
\textbf{12$\boldsymbol\times$} faster than FlashInfer, with the gap
widening monotonically in tree size, depth, and prefix length — the
signature of linear-vs-quadratic scaling against
DeFT and FlashInfer.
\textbf{(2)}~Speed alone does not unlock arbitrarily deep drafts: across
our sweeps mean accepted tokens per step plateau at ${\sim}6.6$ at large
depth, echoing the hidden-state drift ceiling reported by EAGLE-3,
Medusa, and Qwen-class draft analyses --- once attention is no longer
the bottleneck, draft-model fidelity is.
\textbf{(3)}~In FP16 end-to-end on Llama-3.1-8B-Instruct, attention is
only ${\sim}5\%$ of the verification step at the tree sizes EAGLE-3 was
tuned for, precisely because that policy was forced to keep trees small
to dodge FlashInfer's quadratic blow-up; Amdahl's law then caps E2E gains.
\textbf{(4)}~Under W4A16 weight quantization, attention's share of
verification rises and the asymmetry surfaces: \method{} delivers up to
\textbf{1.5$\boldsymbol\times$} E2E with Sequoia at $\numtokens{=}1024$
and up to \textbf{1.8$\boldsymbol\times$} E2E with EAGLE-3 at
$\prefixlen{=}4096$, $\branching{=}32$, $\depth{=}32$ — regimes that
were previously off-limits.
\textbf{(5)}~By making large trees cheap, \method{} reopens a design
space the field had closed for systems reasons: tree-construction
policies (e.g., Sequoia's scaling acceptance curves) can now be revisited
on their statistical merits rather than against a quadratic verifier.
\end{abstract}

\newpage

% =====================================================================
% References
% =====================================================================
\bibliographystyle{plain}
\begin{thebibliography}{20}

\bibitem{leviathan2023fast}
Y.~Leviathan, M.~Kalman, and Y.~Matias.
\newblock Fast inference from transformers via speculative decoding.
\newblock In \emph{ICML}, 2023.

\bibitem{chen2023accelerating}
C.~Chen, S.~Borgeaud, G.~Irving, J.-B.~Lespiau, L.~Sifre, and J.~Jumper.
\newblock Accelerating large language model decoding with speculative sampling.
\newblock \emph{arXiv:2302.01318}, 2023.

\bibitem{li2024eagle2}
Y.~Li, F.~Xia, J.~Zeng, C.~Liu, P.~Peng, and L.~Luo.
\newblock {EAGLE-2}: Faster inference of language models with dynamic draft trees.
\newblock In \emph{EMNLP}, 2024.

\bibitem{li2025eagle3}
Y.~Li, F.~Xia, J.~Zeng, C.~Liu, and L.~Luo.
\newblock {EAGLE-3}: Scaling up speculative decoding against the memory wall.
\newblock \emph{arXiv:2503.01840}, 2025.

\bibitem{cai2024medusa}
T.~Cai, Y.~Li, Z.~Geng, H.~Peng, and T.~Dao.
\newblock Medusa: Simple framework for accelerating {LLM} generation with
  multiple decoding heads.
\newblock In \emph{ICML}, 2024.

\bibitem{chen2024sequoia}
Z.~Chen, X.~Ma, Z.~Zhang, Y.~Lu, and B.~Chen.
\newblock Sequoia: Scalable, robust, and hardware-aware speculative decoding.
\newblock \emph{arXiv:2402.12374}, 2024.

\bibitem{miao2024specinfer}
X.~Miao, G.~Oliaro, Z.~Zhang, X.~Cheng, Z.~Wang, R.~Y.~Wong, Z.~Zhu,
  A.~Alizadeh, Z.~Jia, and I.~Stoica.
\newblock {SpecInfer}: Accelerating generative {LLM} serving with tree-based
  speculative inference and verification.
\newblock In \emph{ASPLOS}, 2024.

\bibitem{dao2022flashattention}
T.~Dao, D.~Y.~Fu, S.~Ermon, A.~Rudra, and C.~R{\'e}.
\newblock Flash{A}ttention: Fast and memory-efficient exact attention with
  {IO}-awareness.
\newblock In \emph{NeurIPS}, 2022.

\bibitem{dao2023flashattention2}
T.~Dao.
\newblock Flash{A}ttention-2: Faster attention with better parallelism and work
  partitioning.
\newblock In \emph{ICLR}, 2024.

\bibitem{ye2025flashinfer}
Z.~Ye, L.~Zheng, Q.~Sun, W.~Wu, J.~Zhang, Z.~Chen, and I.~Stoica.
\newblock {FlashInfer}: Efficient and customizable attention engine for {LLM}
  inference serving.
\newblock \emph{arXiv:2501.01005}, 2025.

\bibitem{yao2024deft}
Y.~Yao, S.~Zuo, S.~Shi, X.~Ning, K.~Wang, Z.~Lu, Y.~Li, and Y.~Wang.
\newblock {DeFT}: Decoding with flash tree-attention for efficient
  tree-structured {LLM} inference.
\newblock \emph{arXiv:2404.00242}, 2024.

\bibitem{tillet2019triton}
P.~Tillet, H.-T.~Kung, and D.~Cox.
\newblock Triton: An intermediate language and compiler for tiled neural
  network computations.
\newblock In \emph{MAPL}, 2019.

\bibitem{zaheer2020bigbird}
M.~Zaheer, G.~Guruganesh, K.~A.~Dubey, J.~Ainslie, C.~Alberti, S.~Ontanon,
  P.~Pham, A.~Ravula, Q.~Wang, L.~Yang, and A.~Ahmed.
\newblock {B}ig{B}ird: Transformers for longer sequences.
\newblock In \emph{NeurIPS}, 2020.

\bibitem{beltagy2020longformer}
I.~Beltagy, M.~E.~Peters, and A.~Cohan.
\newblock Longformer: The long-document transformer.
\newblock \emph{arXiv:2004.05150}, 2020.

\bibitem{kwon2023efficient}
W.~Kwon, Z.~Li, S.~Zhuang, Y.~Sheng, L.~Zheng, C.~H.~Yu, J.~Gonzalez,
  H.~Zhang, and I.~Stoica.
\newblock Efficient memory management for large language model serving with
  {PagedAttention}.
\newblock In \emph{SOSP}, 2023.

\bibitem{sharegpt}
Anon8231489123.
\newblock {ShareGPT\_Vicuna\_unfiltered}: A dataset of human--{GPT}
  conversations.
\newblock \url{https://huggingface.co/datasets/anon8231489123/ShareGPT_Vicuna_unfiltered},
  2023.

\end{thebibliography}

\end{document}
