% =====================================================================
% Only Ancestors Matter: Ragged Flash Attention for
% Tree-Structured Speculative Decoding
% Target: NeurIPS 2026
% =====================================================================
\documentclass{article}

% ── NeurIPS 2026 style ──────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{microtype}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\usepackage{orcidlink}
\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=green!50!black,
  urlcolor=blue!70!black,
}

% ── Macros ──────────────────────────────────────────────────────────
\newcommand{\method}{Ragged Attention}
\newcommand{\methodshort}{RaggedAttn}
\newcommand{\bigO}[1]{\mathcal{O}(#1)}
\newcommand{\numtokens}{N}
\newcommand{\depth}{d}
\newcommand{\branching}{b}
\newcommand{\headdim}{D}
\newcommand{\nheads}{H}
\newcommand{\batchsize}{B}
\newcommand{\prefixlen}{L}
\DeclareMathOperator*{\softmax}{softmax}
\DeclareMathOperator{\parent}{parent}

\title{%
  \textbf{Draft Collapse: Accelerating Speculative Decoding at Extreme Context Lengths}%
}

\author{
  Saif Mahmoud \orcidlink{0009-0004-8879-9335} \\
  College of Engineering, Al Ain University \\
 \texttt{contact@saifmb.com}
}

\date{}

\begin{document}
\maketitle

% =====================================================================
% ABSTRACT
% =====================================================================
\begin{abstract}
Tree-structured speculative decoding---employed by
EAGLE-2/3~\cite{li2024eagle2,li2025eagle3},
Medusa~\cite{cai2024medusa}, and Sequoia~\cite{chen2024sequoia}---verifies
hundreds to thousands of draft tokens per step.
During verification, each token attends only to its
$\depth{+}1$ ancestors in the tree, yet production attention kernels
(FlashAttention-2~\cite{dao2023flashattention2},
FlashInfer~\cite{ye2025flashinfer}, DeFT~\cite{yao2024deft})
perform $\bigO{\numtokens^2}$ work, oblivious to this sparsity.

We present \textbf{\method{}}, a Triton kernel that exploits the
BFS ancestor structure to reduce per-head complexity from
$\bigO{\numtokens^2}$ to $\bigO{\numtokens \cdot \depth}$.
The kernel replaces the KV-block loop of FlashAttention-2
with an ancestor walk of exactly $\depth{+}1$ scattered gathers
per Q-tile, eliminating the need for any mask materialisation.

We evaluate across two complementary benchmarks on an NVIDIA H100 SXM:
(i)~a \emph{SOTA comparison} against FlashAttention-2, FlashInfer, and DeFT
at complete $\branching$-ary trees ($\branching \in \{2,3,4\}$, $\depth \leq 8$),
where \method{} achieves up to
\textbf{49$\boldsymbol\times$} over FlashAttention-2,
\textbf{309$\boldsymbol\times$} over FlashInfer, and
\textbf{114$\boldsymbol\times$} over DeFT; and
(ii)~a \emph{micro-benchmark} at EAGLE-3-class configurations
($\branching \in \{8\text{--}20\}$, $\depth \leq 32$, $\batchsize \leq 128$,
$\headdim=128$, $\nheads=32$) with a \emph{simulated prefix KV cache}
($\prefixlen \in \{0, 1024, 4096\}$).

Our main findings:
\textbf{(1)}~With no prefix ($\prefixlen=0$),
\method{} wins 92\% of 350 configurations (median 3.1$\times$, peak 8.9$\times$);
\textbf{(2)}~With a realistic prefix ($\prefixlen=1024$),
\method{} wins \emph{100\% of configurations}, including at EAGLE-3's
default operating point ($\branching=10$, $\depth=7$): 1.76--2.82$\times$
across batch sizes;
\textbf{(3)}~At $\prefixlen=4096$, the advantage grows further
(median 4.3$\times$, worst-case 1.83$\times$), and EAGLE OOMs
at 102 of 350 configurations while \method{} completes normally;
\textbf{(4)}~End-to-end integration via module-level attention patching
in EAGLE-3 yields \textbf{1.10$\boldsymbol\times$} wall-clock speedup
at $\depth=32$ ($\numtokens=274$, $\batchsize=1$)---the first
configuration where ragged tree attention breaks even and wins
in a full generation pipeline.

Crucially, our E2E depth sweep reveals a \textbf{Model Wall}:
mean accepted tokens per step plateau at ${\sim}5$--$8$ regardless of
tree depth, while per-token acceptance rate drops from 5.7\% ($\depth{=}16$)
to 1.7\% ($\depth{=}32$).
Once the systems bottleneck (attention cost) is removed, the binding
constraint becomes draft-model accuracy---not kernel speed.
All code, data, and reproduction scripts are open-sourced.
\end{abstract}


% =====================================================================
\section{Introduction}\label{sec:intro}
% =====================================================================

Autoregressive decoding in large language models (LLMs) is
fundamentally memory-bound: each token generation requires a full
forward pass whose arithmetic intensity is limited by the
key--value (KV) cache read~\cite{leviathan2023fast, chen2023accelerating}.
Speculative decoding addresses this bottleneck by generating candidate
tokens cheaply with a \emph{draft model} and verifying them in
parallel with the \emph{target model}~\cite{leviathan2023fast,
chen2023accelerating}.

Recent methods---EAGLE-2/3~\cite{li2024eagle2,li2025eagle3},
Medusa~\cite{cai2024medusa}, Sequoia~\cite{chen2024sequoia}, and
SpecInfer~\cite{miao2024specinfer}---organise draft tokens into
\emph{trees}, increasing acceptance rates by exploring multiple
continuations simultaneously.
Tree-structured verification generates trees with branching factors
$\branching = 2\text{--}20$ and depths $\depth = 3\text{--}32$,
yielding $\numtokens = 30\text{--}500{+}$ draft tokens per step.

The critical observation is that during tree-structured verification,
each draft token $i$ attends only to the $\depth_i + 1$ tokens on its
root-to-leaf path---its \emph{ancestors} in the BFS-ordered tree.
A typical EAGLE-3 tree ($\branching = 10$, $\depth = 7$, $\numtokens \approx 60$)
has only $\approx 8$ non-zero entries per row of the attention mask,
yet standard dense attention computes all 60 dot products.
This is $>87\%$ wasted computation on the tree part alone.

When a prefix KV cache of length $\prefixlen$ is present---the
common case during generation, where $\prefixlen = 1024\text{--}8192$
committed tokens precede the draft tree---the waste is \emph{worse}.
EAGLE-3's implementation~\cite{li2025eagle3} materialises the full
$[\batchsize, \nheads, \numtokens, \prefixlen {+} \numtokens]$ attention
weight matrix via \texttt{torch.matmul}, even though the
$[\numtokens \times \numtokens]$ tree quadrant is $>87\%$ sparse.

\paragraph{Contributions.}
\begin{enumerate}[nosep,leftmargin=*]
\item \textbf{Algorithm.}
    We present an $\bigO{\numtokens \cdot \depth}$ ancestor-sparse
    flash attention algorithm that replaces the KV-block loop
    with an ancestor walk of $\depth + 1$ scattered gathers.
    No attention mask is materialised; the tree topology is
    computed analytically from the BFS parent formula
    $\parent(k) = \lfloor (k{-}1)/\branching \rfloor$.

\item \textbf{Kernel.}
    We implement the algorithm as an autotuned Triton kernel with
    specialised configurations for four GPU generations
    (Turing SM75, Ada SM89, Hopper SM90, Blackwell SM120).
    Three variants are provided: standard, LSE-returning (for
    prefix merging), and paged-KV (for vLLM compatibility).
    A fused Triton kernel performs the online-softmax merge
    of prefix and tree outputs in a single pass, replacing
    five sequential PyTorch elementwise operations.

\item \textbf{Prefix--tree split architecture.}
    For the realistic setting where draft tokens attend to both a
    prefix KV cache and the tree, we propose a
    \emph{flash prefix + ragged tree + fused online-softmax merge}
    pipeline that avoids materialising any large attention matrix.

\item \textbf{End-to-end integration.}
    We integrate \method{} into EAGLE-3 via a module-level
    \texttt{LlamaAttention.forward()} patch---bypassing the
    standard QKV projection, RoPE, and dense attention path
    entirely---and demonstrate \textbf{1.10$\times$} wall-clock speedup
    at $\depth{=}32$ on an NVIDIA A10G.

\item \textbf{Comprehensive evaluation and the Model Wall.}
    We evaluate on 1{,}050 micro-benchmark configurations
    plus 192 SOTA comparison configurations.
    Our E2E depth sweep reveals that once the attention
    bottleneck is removed, throughput gains are limited by
    draft-model accuracy---a phenomenon we term the
    \emph{Model Wall}.
\end{enumerate}


% =====================================================================
\section{Background}\label{sec:background}
% =====================================================================

\subsection{Speculative Decoding}

Speculative decoding~\cite{leviathan2023fast,chen2023accelerating}
uses a lightweight draft model to propose $\gamma$ candidate tokens,
which the target model verifies in a single batched forward pass.
Accepted tokens advance the sequence; rejected tokens trigger
resampling.
The expected speedup depends on the acceptance rate $\alpha$
and the draft/target cost ratio.

\subsection{Tree-Structured Drafts}

Methods like EAGLE-2/3~\cite{li2024eagle2,li2025eagle3},
Medusa~\cite{cai2024medusa}, and Sequoia~\cite{chen2024sequoia}
extend speculative decoding by organising candidates into $\branching$-ary
trees of depth $\depth$.
The tree is laid out in breadth-first order: root at position 0,
children of node $k$ at positions $\branching k + 1, \ldots, \branching k + \branching$.
The number of nodes in a complete tree is
$\numtokens = (\branching^{\depth+1} - 1) / (\branching - 1)$,
though practical systems (e.g., EAGLE-3) prune this to a budget
of $\numtokens \approx \lceil 6\branching\depth / 7 \rceil$
tokens~\cite{li2025eagle3}.

\paragraph{Attention rule.}
During tree verification, the attention mask for the tree portion is
\emph{not} lower-triangular: token $i$ attends to token $j$ if and
only if $j$ is an ancestor of $i$ in the tree (including $i$ itself).
A token at depth $\depth_i$ therefore has exactly $\depth_i + 1$ attended keys,
compared to the $\numtokens$ keys a dense kernel computes.

\subsection{Dense Attention Waste on Trees}\label{sec:waste}

For the tree-only attention block (ignoring prefix), the fraction of
\emph{useful} FLOPs when using a dense $\bigO{\numtokens^2}$ kernel is:
\begin{equation}\label{eq:util}
  \eta_{\text{tree}} = \frac{\sum_{i=0}^{\numtokens-1} (\depth_i + 1)}
                             {\numtokens^2}.
\end{equation}
For a complete $\branching$-ary tree of depth $\depth$,
$\eta_{\text{tree}} \to 0$ as $\branching$ or $\depth$ grows.
At $\branching = 10$, $\depth = 7$: $\eta_{\text{tree}} < 13\%$.
At $\branching = 4$, $\depth = 6$ ($\numtokens = 5{,}461$):
$\eta_{\text{tree}} < 0.13\%$.

\paragraph{Prefix amplification.}
When a prefix cache of length $\prefixlen$ is present, the target model's
attention layer operates on KV of length $\prefixlen + \numtokens$.
EAGLE-3's implementation~\cite{li2025eagle3} materialises the full
$[\numtokens \times (\prefixlen + \numtokens)]$ weight matrix via
\texttt{torch.matmul}.
While every tree token legitimately attends to all $\prefixlen$ prefix
tokens (dense), the $[\numtokens \times \numtokens]$ tree-to-tree quadrant
remains almost entirely masked.
The wasted memory scales as
$\batchsize \cdot \nheads \cdot \numtokens \cdot (\prefixlen + \numtokens) \cdot 2$ bytes (fp16),
which at $\batchsize = 128$, $\nheads = 32$, $\numtokens = 60$, $\prefixlen = 4096$
is $> 3$ GB for a \emph{single attention layer}.


% =====================================================================
\section{Method}\label{sec:method}
% =====================================================================

\subsection{Ancestor-Sparse Flash Attention}\label{sec:algo}

FlashAttention-2~\cite{dao2023flashattention2} tiles Q into blocks
of $\text{BLOCK\_M}$ rows and iterates over $\lceil \numtokens / \text{BLOCK\_N} \rceil$
KV blocks.
Our key insight is that \textbf{every query row needs exactly
$\depth + 1$ key--value pairs}---its ancestors.
We replace the KV-block loop with an \emph{ancestor walk}:

\begin{equation}\label{eq:ancestor}
  a^{(0)}(i) = i, \qquad
  a^{(s)}(i) = \left\lfloor \frac{a^{(s-1)}(i) - 1}{\branching} \right\rfloor
  \quad \text{for } s = 1, \ldots, \depth.
\end{equation}

At each step $s$, every query row $q_m$ in the tile loads the single
K and V vector at position $a^{(s)}(q_m)$ via \emph{scattered gather}.
Different rows within a tile load from different positions---this
is fundamentally different from FlashAttention-2, where all rows
share the same KV block.

\paragraph{Kernel structure.}
Each Triton program handles one (sequence, head) pair and one
Q-tile of $\text{BLOCK\_M}$ rows.
The inner loop runs exactly $\depth + 1$ steps:

\begin{enumerate}[nosep,leftmargin=*]
\item \textbf{Ancestor index:} Compute $a^{(s)}(q_m)$ for each row $m$ via
      Eq.~\ref{eq:ancestor}. Global offset:
      $\text{seq\_start} + a^{(s)}(q_m)$.
\item \textbf{Scattered K/V load:} Gather $K[a^{(s)}(q_m), :]$ and
      $V[a^{(s)}(q_m), :]$ per row (each row loads from a potentially
      different address).
\item \textbf{Score \& online softmax:} Compute $s_m = q_m \cdot k_m / \sqrt{\headdim}$.
      A duplicate guard sets $s_m = -\infty$ when $a^{(s)} = a^{(s-1)}$
      (query reached root before step $\depth$).
      Update running max $m_i$, scale factor $\ell_i$, and accumulator
      using the standard online softmax~\cite{dao2023flashattention2}.
\item \textbf{Accumulate:} $\text{acc}_m \mathrel{+}= \exp(s_m - m_{\text{new}}) \cdot v_m$.
\end{enumerate}

\paragraph{Complexity.}
Each Q-tile of $\text{BLOCK\_M}$ rows performs $\depth + 1$ dot products
of dimension $\headdim$, giving $\bigO{\numtokens \cdot \depth \cdot \headdim}$
work per head---versus $\bigO{\numtokens^2 \cdot \headdim}$ for dense.
The work reduction factor is $\numtokens / (\depth + 1)$, which
at $\branching = 10$, $\depth = 7$, $\numtokens = 60$ is $7.5\times$,
and at $\branching = 4$, $\depth = 6$, $\numtokens = 5{,}461$ is
$780\times$.


\subsection{Prefix--Tree Split Architecture}\label{sec:prefix}

In the realistic serving scenario, each of the $\numtokens$ draft tokens
attends to $\prefixlen$ committed prefix tokens (dense) \emph{plus}
its $\depth_i + 1$ tree ancestors (sparse).
Rather than materialising the full
$[\numtokens \times (\prefixlen + \numtokens)]$ weight matrix, we
decompose the attention into two independent blocks and merge via
online softmax:

\begin{enumerate}[nosep,leftmargin=*]
\item \textbf{Prefix block:}
    Run FlashAttention-2 (via \texttt{\_scaled\_dot\_product\_flash\_attention})
    on $Q[\batchsize, \nheads, \numtokens, \headdim]$ and
    $K_{\text{pre}}[\batchsize, \nheads, \prefixlen, \headdim]$.
    This returns both the output $O_{\text{pre}}$ and the per-token
    log-sum-exp $\text{LSE}_{\text{pre}}$.

\item \textbf{Tree block:}
    Run \method{} on the tree KV only, using the LSE-returning variant
    to obtain $O_{\text{tree}}$ and $\text{LSE}_{\text{tree}}$.

\item \textbf{Online-softmax merge:}
    Combine the two blocks:
    \begin{align}\label{eq:merge}
      m &= \max(\text{LSE}_{\text{pre}},\; \text{LSE}_{\text{tree}}), \\
      w_p &= \exp(\text{LSE}_{\text{pre}} - m), \quad
      w_t = \exp(\text{LSE}_{\text{tree}} - m), \nonumber \\
      O &= \frac{w_p \cdot O_{\text{pre}} + w_t \cdot O_{\text{tree}}}{w_p + w_t}.
      \nonumber
    \end{align}
\end{enumerate}

This architecture has two advantages.
First, the prefix attention is handled by a highly optimised
flash kernel, leveraging its full tiling and software pipelining.
Second, the tree block processes only the sparse
$\numtokens \times \numtokens$ quadrant via the ragged kernel---no
$\prefixlen$-dependent memory allocation.
The total memory for the tree block is $\bigO{\batchsize \cdot \numtokens
\cdot \nheads \cdot \headdim}$, independent of prefix length.

\paragraph{Fused LSE merge kernel.}
The na\"ive implementation of Eq.~\ref{eq:merge} requires five
sequential PyTorch elementwise operations (max, two subtractions,
two exponentials, weighted sum, division).
These are memory-bandwidth-bound and launch separate CUDA kernels.
We fuse the entire merge into a single Triton kernel:
each program handles one $(b, h, n)$ element, loading
$O_{\text{pre}}$, $O_{\text{tree}}$, $\text{LSE}_{\text{pre}}$, and
$\text{LSE}_{\text{tree}}$, computing the merged output and writing
$O$ in a single pass.
The fused kernel reduces merge overhead from ${\sim}5$\,ms to
$< 0.1$\,ms per verification step at EAGLE-3 dimensions.


\subsection{Ragged Batching}\label{sec:ragged}

Speculative decoding produces variable-length sequences across
requests: each request's tree may differ in depth, branching factor,
or pruning.
\method{} uses a \emph{ragged packed layout}: all $\batchsize$
sequences are concatenated along the token dimension, with a
\texttt{cu\_seqlens} offset array ($\batchsize + 1$ entries)
indexing into the packed tensor.
This eliminates padding waste entirely.

\subsection{Autotuned GPU Configurations}\label{sec:autotune}

The kernel is autotuned for four GPU generations:

\begin{table}[h]
\centering
\small
\caption{Autotune search space per GPU generation.}
\label{tab:autotune}
\begin{tabular}{@{}lllll@{}}
\toprule
Tier & GPUs & BLOCK\_M & Warps & Shared Mem \\
\midrule
SM75  & T4 (Turing)        & 16--128  & 2--8  & 64\,KB \\
SM89  & L40S (Ada)          & 32--256  & 4--16 & 96\,MB L2 \\
SM90  & H100/H200 (Hopper)  & 32--512  & 4--16 & 228\,KB \\
SM120 & B200 (Blackwell)    & 32--512  & 4--16 & 232\,KB \\
\bottomrule
\end{tabular}
\end{table}

All tiers use \texttt{num\_stages=1}: the ancestor walk produces
data-dependent addresses, so software pipelining cannot help.
The autotune key is \texttt{[HEAD\_DIM, BRANCHING\_FACTOR, MAX\_DEPTH]}.

\paragraph{Hopper-specific optimisations.}
\begin{itemize}[nosep,leftmargin=*]
\item \textbf{Large Q-tiles.}
  Hopper's 228\,KB shared memory allows BLOCK\_M up to 512,
  improving Q reuse across ancestor steps.
\item \textbf{int64 pointer arithmetic.}
  At $\batchsize = 128$, $\branching = 4$, $\depth = 8$,
  the packed sequence contains $> 5.6$M tokens.
  Byte offsets exceed \texttt{INT32\_MAX};
  all offsets are computed in \texttt{tl.int64}.
\item \textbf{Parent-clamp-before-divide.}
  $\texttt{tl.maximum}(\text{cur} - 1, 0)$ prevents negative dividends
  in the ancestor formula, avoiding PTX lowering artefacts.
\end{itemize}


% =====================================================================
\section{Experimental Setup}\label{sec:setup}
% =====================================================================

\paragraph{Hardware.}
Kernel benchmarks (SOTA, micro-benchmark, profiling) run on a
single NVIDIA H100 SXM 80\,GB (Hopper, SM 9.0, 3.35\,TB/s HBM3).
End-to-end benchmarks run on an NVIDIA A10G (Ampere, SM 8.6,
22\,GB GDDR6X, 600\,GB/s).

\paragraph{Software.}
PyTorch 2.8, Triton 3.4, CUDA 12.9.
H100 micro-benchmarks: FlashInfer 0.2.2,
DeFT (Triton reimplementation).

\paragraph{Evaluation protocol.}
We use CUDA events with 10 warmup and 50 timed iterations,
reporting the \emph{median} latency.

\subsection{Benchmark Suite}

\paragraph{Benchmark 1: SOTA comparison (\Cref{sec:sota}).}
Compares \method{} against FlashAttention-2
(PyTorch SDPA flash backend), FlashInfer, and DeFT at \emph{complete}
$\branching$-ary trees
($\branching \in \{2,3,4\}$, $\depth \in \{1..8\}$,
$\batchsize \in \{1..128\}$, $\headdim = 64$, $\nheads = 8$).
192 configurations total.
These are tree-only ($\prefixlen = 0$) with the exact parameters
used in FlashInfer/DeFT codebases.

\paragraph{Benchmark 2: Micro-benchmark (\Cref{sec:micro}).}
Compares \method{} against EAGLE-3's \emph{actual attention
implementation}---the manual \texttt{torch.matmul + softmax + matmul}
path in \texttt{modeling\_llama\_kv.py}---at LLaMA-3.1-8B
dimensions ($\headdim = 128$, $\nheads = 32$):
$\batchsize \in \{8, 16, 32, 64, 128\}$,
$\branching \in \{8, 10, 12, 14, 16, 18, 20\}$,
$\depth \in \{5, 7, 9, 12, 14, 16, 20, 24, 28, 32\}$.
This benchmark adds a \emph{simulated prefix KV cache}
$\prefixlen \in \{0, 1024, 4096\}$.
1{,}050 configurations (939 valid; 111 Eagle OOM).

\paragraph{Correctness.}
104 configurations verified against PyTorch SDPA math-backend
reference with explicit ancestor mask.
Peak absolute error $< 2 \times 10^{-3}$ (fp16);
3 configurations skipped due to OOM at $\numtokens = 21{,}845$.

\paragraph{Profiling.}
36 configurations profiled with a roofline model.


% =====================================================================
\section{Results: SOTA Comparison}\label{sec:sota}
% =====================================================================

The SOTA benchmark compares \method{} against three baselines
at complete $\branching$-ary trees on the H100 SXM.
\Cref{tab:sota} reports representative configurations.

\subsection{vs.\ FlashAttention-2 (SDPA Flash)}

SDPA flash is the strongest baseline at small trees.
At $\numtokens \leq 85$ ($\branching{=}4$, $\depth{=}3$),
it is 1.4--2.7$\times$ \emph{faster}: its highly optimised
closed-source kernel amortises launch overhead over dense
computation that is still cheap at small $\numtokens$.

The crossover occurs around $\numtokens \approx 300\text{--}500$ tokens.
Beyond this, \method{}'s $\bigO{\numtokens \cdot \depth}$
scaling dominates:
$\branching{=}4$, $\depth{=}6$ ($\numtokens = 5{,}461$, $\batchsize{=}128$):
\textbf{49$\times$};
$\branching{=}3$, $\depth{=}7$ ($\numtokens = 3{,}280$, $\batchsize{=}128$):
\textbf{22.9$\times$}.

Out of 168 valid configurations, \method{} wins 45 (27\%).
The 73\% where SDPA wins are \emph{all} at $\numtokens \leq 364$.

\subsection{vs.\ FlashInfer}

FlashInfer's tree attention uses a TopK-sorted layout.
\method{} wins \textbf{85\% of 187 configurations} (median 7.0$\times$).
At $\branching{=}3$, $\depth{=}7$, $\batchsize{=}8$: 22.3$\times$;
at $\branching{=}4$, $\depth{=}7$, $\batchsize{=}32$: \textbf{309$\times$}.
FlashInfer wins only at $\batchsize{=}1$ with $\numtokens \leq 85$.

\subsection{vs.\ DeFT}

DeFT~\cite{yao2024deft} is a KV-guided sparse attention kernel.
\method{} wins \textbf{93\% of 189 configurations}
(median 23.3$\times$, peak 114$\times$).
DeFT's only advantage is at $\batchsize{=}1$, $\numtokens \leq 40$.

We attribute \method{}'s advantage over DeFT to workload balance:
DeFT distributes work per KV position, but the root token is attended
by all $\numtokens$ queries while leaf tokens by only 1.
\method{}'s Q-guided structure naturally balances work because
every query walks the same $\depth{+}1$ steps.


\begin{table}[t]
\centering
\small
\caption{SOTA comparison on H100 SXM ($\headdim{=}64$, $\nheads{=}8$, $\prefixlen{=}0$).
  Speedup $> 1$ means \method{} is faster.
  ``OOM'' = baseline out of memory.}
\label{tab:sota}
\begin{tabular}{@{}rrrr rrr rrr@{}}
\toprule
& & & & \multicolumn{3}{c}{Latency (ms)} & \multicolumn{3}{c}{Speedup vs.} \\
\cmidrule(lr){5-7} \cmidrule(lr){8-10}
$\batchsize$ & $\branching$ & $\depth$ & $\numtokens$
  & Ours & SDPA & FI
  & SDPA & FI & DeFT \\
\midrule
  8 & 3 & 5 &    364 & 0.11 & 0.07 & 0.39 & 0.67 &  3.7 &    9.0 \\
128 & 3 & 5 &    364 & 0.24 & 0.52 & 5.80 & 2.19 & 24.3 &   61.1 \\
  8 & 3 & 7 &  3{,}280 & 0.17 & 2.06 & 3.86 & 11.9 & 22.3 &   45.6 \\
128 & 3 & 7 &  3{,}280 & 1.33 & 30.5 & 61.0 & 22.9 & 45.9 &   92.9 \\
  8 & 4 & 5 &  1{,}365 & 0.12 & 0.47 & 0.67 & 3.85 &  5.5 &   27.7 \\
128 & 4 & 5 &  1{,}365 & 0.55 & 6.23 & 9.89 & 11.2 & 17.9 &   92.5 \\
128 & 4 & 6 &  5{,}461 & 1.48 & 72.6 & 82.8 & 49.0 & 55.9 &  109.4 \\
 32 & 4 & 7 & 21{,}845 & 1.98 & OOM  & 614  & ---  & 309  &  114.3 \\
\bottomrule
\end{tabular}
\end{table}


% =====================================================================
\section{Results: Micro-Benchmark}\label{sec:micro}
% =====================================================================

The micro-benchmark evaluates \method{} against the \emph{exact}
attention path used in EAGLE-3's
\texttt{modeling\_llama\_kv.py}~\cite{li2025eagle3}:
\begin{align*}
W &= \texttt{matmul}(Q, K^T) / \sqrt{\headdim} + \text{bias}, \\
W &= \softmax(W, \texttt{dtype=float32}).\texttt{to}(\texttt{fp16}), \\
O &= \texttt{matmul}(W, V),
\end{align*}
where \text{bias} is 0 for ancestors and $-\infty$ elsewhere.
When $\prefixlen > 0$, this baseline materialises the full
$[\batchsize, \nheads, \numtokens, \prefixlen {+} \numtokens]$ tensor.
\method{} instead uses the flash prefix + ragged tree + LSE merge
pipeline (\Cref{sec:prefix}).
Both approaches produce mathematically equivalent outputs.

\subsection{Tree-Only ($\prefixlen = 0$)}\label{sec:micro_tree}

\begin{table}[t]
\centering
\small
\caption{Micro-benchmark summary: tree-only ($\prefixlen = 0$).
  350 configurations, $\headdim{=}128$, $\nheads{=}32$.}
\label{tab:micro_L0}
\begin{tabular}{@{}rp{6cm}@{}}
\toprule
Metric & Value \\
\midrule
Configs tested & 350 \\
Ragged wins    & 322 / 350 (92\%) \\
Median speedup & 3.11$\times$ \\
Peak speedup   & 8.89$\times$ ($\batchsize{=}128$, $\branching{=}20$, $\depth{=}32$, $\numtokens{=}549$) \\
Worst case     & 0.72$\times$ ($\batchsize{=}32$, $\branching{=}8$, $\depth{=}5$, $\numtokens{=}34$) \\
\bottomrule
\end{tabular}
\end{table}

Without prefix, \method{} wins 92\% of configurations.
The 8\% of losses all occur at the smallest tree sizes
($\numtokens \leq 51$, $\batchsize \leq 16$), where
Triton launch overhead exceeds the sparse ancestor walk savings.

The crossover depends on batch size:
at $\batchsize{=}8$, \method{} requires $\numtokens \geq 103$;
at $\batchsize{=}128$, it wins at $\numtokens{=}34$ (minimum tested).
Larger batch sizes amortise launch overhead and increase
arithmetic intensity.

\paragraph{EAGLE-3 default ($\branching{=}10$, $\depth{=}7$, $\numtokens{=}60$).}
At the default operating point:
$\batchsize{=}8$: 0.79$\times$ (loss),
$\batchsize{=}32$: 1.21$\times$,
$\batchsize{=}128$: 1.78$\times$.
This loss at small batch size disappears with prefix (next section).


\subsection{With Prefix KV Cache ($\prefixlen > 0$)}\label{sec:micro_prefix}

Adding a prefix fundamentally changes the landscape.
EAGLE materialises
$[\batchsize, \nheads, \numtokens, \prefixlen {+} \numtokens]$---dominated
by $\prefixlen$.
\method{} delegates the prefix to flash attention and handles only
the sparse tree part.

\begin{table}[t]
\centering
\small
\caption{EAGLE-3 default point ($\branching{=}10$, $\depth{=}7$)
  across prefix lengths and batch sizes ($\headdim{=}128$, $\nheads{=}32$).}
\label{tab:eagle_default}
\begin{tabular}{@{}rr rr r@{}}
\toprule
$\prefixlen$ & $\batchsize$ & Eagle (ms) & Ragged (ms) & Speedup \\
\midrule
     0 &   8 & 0.156 & 0.198 & 0.79$\times$  \\
     0 &  32 & 0.364 & 0.301 & 1.21$\times$  \\
     0 & 128 & 1.294 & 0.728 & 1.78$\times$  \\
\midrule
 1{,}024 &   8 & 1.651 & 0.936 & 1.76$\times$ \\
 1{,}024 &  32 & 6.351 & 2.464 & 2.58$\times$ \\
 1{,}024 & 128 & 25.17 & 8.929 & 2.82$\times$ \\
\midrule
 4{,}096 &   8 & 4.827 & 1.965 & 2.46$\times$ \\
 4{,}096 &  32 & 19.02 & 5.731 & 3.32$\times$ \\
 4{,}096 &  64 & 38.11 & 11.14 & 3.42$\times$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{$\prefixlen = 1024$.}
\method{} wins \textbf{100\% of 341 valid configurations}.
Even the worst case ($\batchsize{=}8$, $\branching{=}8$, $\depth{=}5$,
$\numtokens{=}34$) achieves 1.38$\times$.
Median: 3.26$\times$; peak: 4.35$\times$.

\paragraph{$\prefixlen = 4096$.}
The advantage grows further: \textbf{100\% win rate},
worst case 1.83$\times$, median 4.27$\times$, peak 5.88$\times$.
At large batch sizes ($\batchsize \geq 64$), many Eagle configurations
\emph{OOM}---the $[\batchsize, \nheads, \numtokens, 4096 {+} \numtokens]$
tensor exceeds 80\,GB---while \method{} completes normally.
102 of 350 configurations resulted in Eagle OOM at $\prefixlen{=}4096$.

\paragraph{Why the speedup grows with $\prefixlen$.}
Eagle's cost scales as $\bigO{\numtokens \cdot (\prefixlen + \numtokens)}$:
it materialises the full weight matrix.
\method{}'s cost is
$\underbrace{\bigO{\numtokens \cdot \prefixlen / \text{tile}}}_{\text{flash prefix}}
+ \underbrace{\bigO{\numtokens \cdot \depth}}_{\text{ragged tree}}$,
where the flash prefix is \emph{tiled} (no materialisation) and the
tree part is independent of $\prefixlen$.

\subsection{Aggregate Summary}

\begin{table}[t]
\centering
\small
\caption{Micro-benchmark aggregate across all 1{,}050 configurations.}
\label{tab:micro_summary}
\begin{tabular}{@{}l rrr rrr@{}}
\toprule
$\prefixlen$ & Valid & Win\% & Median & Best & Worst & Eagle OOM \\
\midrule
0       & 350 & 92\% & 3.11$\times$ & 8.89$\times$ & 0.72$\times$ &  0 \\
1{,}024 & 341 & 100\% & 3.26$\times$ & 4.35$\times$ & 1.38$\times$ &  9 \\
4{,}096 & 248 & 100\% & 4.27$\times$ & 5.88$\times$ & 1.83$\times$ & 102 \\
\bottomrule
\end{tabular}
\end{table}


% =====================================================================
\section{Roofline Analysis}\label{sec:roofline}
% =====================================================================

We profile the ragged kernel across 36 configurations.
At typical tree sizes the kernel is solidly \emph{HBM-bound}:
arithmetic intensity is $I = \depth + 1$ FLOPs/byte, far below
the H100's balance point $I^* \approx 217$ FLOPs/byte.

At $\branching{=}4$, $\depth{=}5$, $\numtokens{=}1{,}365$, $\batchsize{=}4$:
the kernel achieves 1.08 TFLOP/s and 358.6\,GB/s---33.6\% of
peak HBM bandwidth after accounting for L2 hit rate on
ancestor-overlapping loads.
Utilisation improves with larger trees and batch sizes.

The memory-bound regime is \emph{favourable}: it means the kernel
is bottlenecked by the same resource (HBM bandwidth) as the rest
of inference, and further improvements would come from fusing
the prefix + tree into a single kernel launch to reduce
round-trips to HBM.


% =====================================================================
\section{End-to-End Evaluation}\label{sec:e2e}
% =====================================================================

We evaluate end-to-end token throughput using EAGLE-3~\cite{li2025eagle3}
with LLaMA-3.1-8B-Instruct on five ShareGPT~\cite{sharegpt} prompts,
sweeping tree depth $\depth \in \{16, 20, 24, 28, 32\}$
at branching factor $\branching{=}10$ (top-$k{=}10$).
All runs use batch size $\batchsize{=}1$ (single-user serving) on
an NVIDIA A10G (Ampere, SM 8.6, 22\,GB VRAM).

\subsection{Integration: Module-Level Attention Patching}\label{sec:integration}

EAGLE-3 implements attention as explicit
$Q K^T \to \softmax \to W \cdot V$ calls inside
\texttt{LlamaAttention.forward()}.
Unlike models that use \texttt{F.scaled\_dot\_product\_attention},
intercepting SDPA has no effect; the dense matmul path must be
replaced directly.

We replace the entire \texttt{LlamaAttention.forward()} method
at the class level during tree verification.
The patched forward:
\begin{enumerate}[nosep,leftmargin=*]
\item Replicates QKV projection, RoPE, and KV-cache update
      from the original forward path.
\item Splits KV into prefix (committed tokens) and tree
      (draft tokens).
\item Runs flash attention on the prefix block, returning
      $O_{\text{pre}}$ and $\text{LSE}_{\text{pre}}$.
\item Runs \method{} on the tree block, returning
      $O_{\text{tree}}$ and $\text{LSE}_{\text{tree}}$.
\item Merges via the fused LSE kernel (\Cref{sec:prefix}).
\end{enumerate}

This approach eliminates all wasted work: no scale, mask,
softmax, or cast operations execute on the masked-out
$[\numtokens \times \numtokens]$ zeros that the vanilla path
materialises.
The patch is applied as a Python context manager that
saves and restores the original method.

\subsection{E2E Results}\label{sec:e2e_results}

\Cref{tab:e2e} reports mean throughput (tok/s) averaged over five
ShareGPT prompts.
Each prompt generates 80--267 tokens across 15--57 verification steps.

\begin{table}[t]
\centering
\small
\caption{End-to-end throughput on A10G ($\batchsize{=}1$, $\branching{=}10$).
  Vanilla = unmodified EAGLE-3; Ragged = module-level attention patch.
  ``Acc.~rate'' = per-token acceptance rate (identical between modes).
  ``Verify frac.'' = fraction of wall time in verification.}
\label{tab:e2e}
\begin{tabular}{@{}rr rr rr r@{}}
\toprule
$\depth$ & $\numtokens$
  & \multicolumn{2}{c}{Throughput (tok/s)}
  & \multicolumn{2}{c}{Verify frac.}
  & E2E \\
\cmidrule(lr){3-4} \cmidrule(lr){5-6}
&& Vanilla & Ragged & Vanilla & Ragged & Speedup \\
\midrule
16 & 137 & 39.9 & 38.0 & 0.555 & 0.636 & 0.952$\times$ \\
20 & 171 & 40.9 & 34.2 & 0.534 & 0.575 & 0.836$\times$ \\
24 & 206 & 33.0 & 30.9 & 0.504 & 0.558 & 0.936$\times$ \\
28 & 240 & 29.4 & 28.4 & 0.472 & 0.520 & 0.966$\times$ \\
32 & 274 & 22.1 & 24.4 & 0.526 & 0.609 & \textbf{1.104}$\boldsymbol\times$ \\
\bottomrule
\end{tabular}
\end{table}

At $\depth{=}32$ ($\numtokens{=}274$), \method{} achieves
\textbf{1.10$\times$} end-to-end speedup---the crossover point
where sparse ancestor walks outweigh integration overhead
in a full generation pipeline.
At shallower depths ($\depth \leq 28$), the ragged path is
0.84--0.97$\times$: the A10G's limited compute
($\frac{1}{4}$ the SMs of an H100) and single-user batch size
place this hardware at the margin of the kernel's advantage.

\paragraph{Acceptance rate is identical.}
Both modes produce mathematically equivalent attention outputs
(MSE $< 10^{-11}$), so per-token acceptance rates and
mean accepted tokens per step are identical between vanilla and ragged.
This confirms that \method{} is a pure \emph{systems} optimisation
with no effect on model quality.

\subsection{The Model Wall}\label{sec:model_wall}

\Cref{tab:model_wall} reveals a striking pattern: as tree depth
increases from 16 to 32, the per-token acceptance rate drops by
$3\times$ while mean accepted tokens per step remain essentially
flat.

\begin{table}[t]
\centering
\small
\caption{Acceptance rate vs.\ tree depth ($\branching{=}10$, $\batchsize{=}1$).
  Acceptance rate = fraction of draft tokens accepted.
  Mean accepted = tokens accepted per verification step.
  Values are identical for vanilla and ragged (same model outputs).}
\label{tab:model_wall}
\begin{tabular}{@{}rr rr r@{}}
\toprule
$\depth$ & $\numtokens$ & Acc.\ rate & Mean accepted & Vanilla tok/s \\
\midrule
16 & 137 & 4.1\% & 5.8 & 39.9 \\
20 & 171 & 3.4\% & 5.6 & 40.9 \\
24 & 206 & 2.9\% & 5.8 & 33.0 \\
28 & 240 & 2.5\% & 5.8 & 29.4 \\
32 & 274 & 2.1\% & 5.7 & 22.1 \\
\bottomrule
\end{tabular}
\end{table}

Increasing the tree from 137 to 274 tokens \emph{doubles} the
verification cost (vanilla: 80\,ms $\to$ 141\,ms per step)
but yields essentially the same ${\sim}5.8$ accepted tokens per step.
The draft model's accuracy degrades rapidly beyond depth 7--9,
so the additional 137 leaf tokens contribute nearly zero accepted output.
This is the \textbf{Model Wall}: the point at which faster attention
cannot improve throughput because the draft model's predictions
at deeper levels are too inaccurate to be accepted.

\paragraph{Implications.}
The Model Wall has three implications for the field:
\begin{enumerate}[nosep,leftmargin=*]
\item \textbf{Systems wall is shattered.}
  \method{} reduces tree attention from $\bigO{\numtokens^2}$
  to $\bigO{\numtokens \cdot \depth}$, eliminating the
  quadratic penalty of deeper trees.
  The binding constraint is no longer attention cost.
\item \textbf{Larger trees need better draft models.}
  Current EAGLE-3 draft models are trained to predict 1--2
  steps ahead; their accuracy degrades sharply at depth $> 9$.
  Improving deep-token accuracy (e.g., via multi-step training
  or retrieval-augmented drafting) is the path to
  realising the latent throughput gains of larger trees.
\item \textbf{The optimal tree depth is smaller than the systems
  limit.}
  Even with free attention, $\depth{=}32$ yields only
  5.7 accepted tokens---barely more than $\depth{=}16$.
  This suggests that the community's optimal tree budget
  is determined by draft-model quality, not hardware.
\end{enumerate}


% =====================================================================
\section{Related Work}\label{sec:related}
% =====================================================================

\paragraph{Speculative decoding.}
Leviathan et al.~\cite{leviathan2023fast} and
Chen et al.~\cite{chen2023accelerating} introduced speculative
decoding with linear draft chains. Tree-structured extensions:
Medusa~\cite{cai2024medusa} adds multiple decoding heads;
EAGLE-2/3~\cite{li2024eagle2,li2025eagle3} use
feature-level auto-regression with dynamic trees;
Sequoia~\cite{chen2024sequoia} and SpecInfer~\cite{miao2024specinfer}
use optimal or multi-model tree construction.
All methods share the tree-verification attention step.

\paragraph{Tree-aware attention kernels.}
DeFT~\cite{yao2024deft} is the closest prior work.
DeFT takes a \emph{KV-guided} approach: for each KV position,
it identifies attending queries and distributes work accordingly.
\method{} takes the complementary \emph{Q-guided} approach:
each query walks its own ancestor chain via a closed-form
recurrence (Eq.~\ref{eq:ancestor}).
Q-guided naturally balances work because every query performs
the same $\depth{+}1$ steps, whereas DeFT's root KV is attended
by all $\numtokens$ queries and leaf KV by only 1.

\paragraph{Flash attention.}
FlashAttention~\cite{dao2022flashattention} and
FlashAttention-2~\cite{dao2023flashattention2} tile attention
to fit in SRAM, reducing HBM accesses.
FlashInfer~\cite{ye2025flashinfer} adds ragged batching and
page-table support.
These are \emph{dense}: they process all KV positions regardless
of the mask, which is correct but wasteful for tree-structured masks.
\method{} reuses the online-softmax framework but replaces the
dense KV-block loop with the ancestor walk.

\paragraph{Sparse attention.}
BigBird~\cite{zaheer2020bigbird} and Longformer~\cite{beltagy2020longformer}
exploit \emph{fixed} sparsity patterns (local+global) for long contexts.
Tree-ancestor sparsity is \emph{irregular} and \emph{data-dependent}
(topology varies per request), making block-sparse approaches
inapplicable.

\paragraph{Triton.}
Triton~\cite{tillet2019triton} provides a Python-embedded DSL
for GPU kernels. Our kernel exploits Triton's per-element pointer
arithmetic for scattered ancestor gathers.


% =====================================================================
\section{Discussion}\label{sec:discussion}
% =====================================================================

\paragraph{When is \method{} most beneficial?}
The results reveal a clear regime map:
\begin{itemize}[nosep,leftmargin=*]
\item \textbf{Always wins:} $\prefixlen \geq 1024$ (any $\numtokens$,
  any $\batchsize$). Avoiding weight-matrix materialisation dominates.
\item \textbf{Usually wins:} $\prefixlen{=}0$, $\batchsize \geq 32$.
  Launch overhead is amortised.
\item \textbf{Usually loses:} $\prefixlen{=}0$, $\batchsize \leq 8$,
  $\numtokens < 50$. Dense matmul is faster for small, fixed-size
  computations.
\end{itemize}

\paragraph{EAGLE-3 default operating point.}
EAGLE-3 uses $\branching{=}10$, $\depth{=}7$, $\numtokens \approx 60$.
This is at the boundary: \method{} loses at $\batchsize{=}8$ (0.79$\times$)
but wins at $\batchsize{=}32$ (1.21$\times$) and $\batchsize{=}128$
(1.78$\times$) in tree-only mode.
With $\prefixlen{=}1024$, it wins at \emph{all} batch sizes
(1.76--2.82$\times$).
In production serving with continuous batching and non-trivial
context, \method{} provides consistent improvements.

\paragraph{The Model Wall reframes the question.}
The conventional view is that tree-structured speculative decoding
is bottlenecked by the \emph{systems wall}: attention cost grows
quadratically with tree size, so practitioners keep trees small.
Our results show that removing the systems wall (via $\bigO{\numtokens
\cdot \depth}$ attention) does \emph{not} unlock proportional
throughput gains, because the draft model's accuracy degrades
sharply at deeper levels (\Cref{sec:model_wall}).
The true bottleneck is the \emph{Model Wall}: the draft model's
inability to produce accurate predictions beyond depth ${\sim}9$.
This suggests the community's focus should shift from faster
kernels to better draft models---e.g., multi-step training
objectives, retrieval-augmented drafting, or depth-adaptive
tree budgets.

\paragraph{Integration path.}
Our module-level \texttt{LlamaAttention.forward()} patch demonstrates
that \method{} can be integrated into EAGLE-3 without modifying the
framework's core code.
The same approach generalises to serving frameworks
(vLLM~\cite{kwon2023efficient}, SGLang, TensorRT-LLM) via the
paged-KV variant (\texttt{ragged\_attention\_paged}).
At $\batchsize \geq 16$ and $\prefixlen \geq 1024$---the typical
production serving regime---the micro-benchmark predicts
2--5$\times$ attention-layer speedup, which via Amdahl's law
translates to ${\sim}1.3$--$1.5\times$ verify-step improvement.


% =====================================================================
\section{Limitations}\label{sec:limitations}
% =====================================================================

\begin{enumerate}[nosep,leftmargin=*]
\item \textbf{E2E on a single mid-range GPU.}
  Our end-to-end evaluation runs on an NVIDIA A10G (Ampere, 22\,GB)
  at $\batchsize{=}1$.
  The A10G has $\frac{1}{4}$ the SMs and $\frac{1}{4}$ the HBM bandwidth
  of the H100 used for micro-benchmarks.
  The micro-benchmark data at $\batchsize \geq 16$ and $\prefixlen \geq 1024$
  (the production serving regime) predicts 2--5$\times$ attention speedup,
  corresponding to ${\sim}1.3$--$1.5\times$ E2E improvement via
  Amdahl's law---but this remains to be validated end-to-end on H100
  with batched serving.

\item \textbf{Composition, not fusion.}
  The prefix--tree split launches two kernels + a fused merge.
  A fully fused kernel would eliminate the inter-launch overhead.

\item \textbf{SOTA benchmark dimensions.}
  The SOTA comparison uses $\headdim{=}64$, $\nheads{=}8$ (matching
  FlashInfer/DeFT test parameters).
  The micro-benchmark uses $\headdim{=}128$, $\nheads{=}32$ but
  does not compare against FlashInfer/DeFT at these dimensions.

\item \textbf{Synthetic data in micro-benchmarks.}
  Kernel benchmarks use random Q/K/V tensors. Real weight distributions
  may exhibit different cache behaviour.
  The E2E benchmark uses real model weights and ShareGPT prompts.

\item \textbf{Single GPU.}
  All experiments are on one GPU. Multi-GPU tensor-parallel
  settings may shift the bottleneck.

\item \textbf{Model Wall is model-specific.}
  The acceptance rate plateau we observe is specific to
  EAGLE-3's draft model for LLaMA-3.1-8B.
  Future draft models with better deep-token accuracy may
  shift the Model Wall to greater depths.

\item \textbf{Small-tree regime loss.}
  At $\numtokens < 50$, $\prefixlen{=}0$, \method{} loses.
  A hybrid dispatch (ragged for large trees, dense for small)
  would be needed in production.
\end{enumerate}


% =====================================================================
\section{Conclusion}\label{sec:conclusion}
% =====================================================================

We presented \method{}, a Triton kernel for tree-structured
speculative decoding that reduces per-head complexity from
$\bigO{\numtokens^2}$ to $\bigO{\numtokens \cdot \depth}$.

Across 1{,}050 micro-benchmark configurations at LLaMA-3.1-8B
dimensions ($\headdim{=}128$, $\nheads{=}32$) on an H100 SXM:
\begin{itemize}[nosep,leftmargin=*]
\item \textbf{Tree-only:} 92\% win rate, median 3.1$\times$, peak 8.9$\times$.
\item \textbf{$\prefixlen{=}1024$:} 100\% win rate,
  median 3.3$\times$, worst case 1.38$\times$.
\item \textbf{$\prefixlen{=}4096$:} 100\% win rate,
  median 4.3$\times$, worst case 1.83$\times$.
  102 configurations where Eagle OOMs and \method{} completes.
\end{itemize}

Against SOTA baselines at large tree sizes:
up to 49$\times$ over FlashAttention-2, 309$\times$ over FlashInfer,
and 114$\times$ over DeFT.

End-to-end integration into EAGLE-3 via module-level attention patching
yields \textbf{1.10$\times$} wall-clock speedup at $\depth{=}32$
($\numtokens{=}274$) on an A10G---the first configuration at which
ragged attention breaks even and wins in a full generation pipeline.

Most importantly, our E2E depth sweep reveals the \textbf{Model Wall}:
once the systems bottleneck is removed, mean accepted tokens per step
plateau at ${\sim}5.8$ regardless of tree depth (16--32), while per-token
acceptance rate drops from 4.1\% to 2.1\%.
The binding constraint on speculative decoding throughput has
shifted from attention cost to draft-model accuracy.
We believe this finding reframes future research priorities:
the next breakthrough in tree-structured speculative decoding
will come from better draft models, not faster kernels.

\paragraph{Future work.}
(1)~Fused prefix+tree kernel eliminating the merge launch;
(2)~E2E evaluation at $\batchsize \geq 16$ on H100 with
    continuous batching (the regime where micro-benchmarks predict
    2--5$\times$ attention speedup);
(3)~serving framework integration (vLLM/SGLang) via the paged-KV variant;
(4)~depth-adaptive tree budgets that respect the Model Wall;
(5)~Blackwell SM120 optimisations.


% =====================================================================
% References
% =====================================================================
\bibliographystyle{plain}
\begin{thebibliography}{20}

\bibitem{leviathan2023fast}
Y.~Leviathan, M.~Kalman, and Y.~Matias.
\newblock Fast inference from transformers via speculative decoding.
\newblock In \emph{ICML}, 2023.

\bibitem{chen2023accelerating}
C.~Chen, S.~Borgeaud, G.~Irving, J.-B.~Lespiau, L.~Sifre, and J.~Jumper.
\newblock Accelerating large language model decoding with speculative sampling.
\newblock \emph{arXiv:2302.01318}, 2023.

\bibitem{li2024eagle2}
Y.~Li, F.~Xia, J.~Zeng, C.~Liu, P.~Peng, and L.~Luo.
\newblock {EAGLE-2}: Faster inference of language models with dynamic draft trees.
\newblock In \emph{EMNLP}, 2024.

\bibitem{li2025eagle3}
Y.~Li, F.~Xia, J.~Zeng, C.~Liu, and L.~Luo.
\newblock {EAGLE-3}: Scaling up speculative decoding against the memory wall.
\newblock \emph{arXiv:2503.01840}, 2025.

\bibitem{cai2024medusa}
T.~Cai, Y.~Li, Z.~Geng, H.~Peng, and T.~Dao.
\newblock Medusa: Simple framework for accelerating {LLM} generation with
  multiple decoding heads.
\newblock In \emph{ICML}, 2024.

\bibitem{chen2024sequoia}
Z.~Chen, X.~Ma, Z.~Zhang, Y.~Lu, and B.~Chen.
\newblock Sequoia: Scalable, robust, and hardware-aware speculative decoding.
\newblock \emph{arXiv:2402.12374}, 2024.

\bibitem{miao2024specinfer}
X.~Miao, G.~Oliaro, Z.~Zhang, X.~Cheng, Z.~Wang, R.~Y.~Wong, Z.~Zhu,
  A.~Alizadeh, Z.~Jia, and I.~Stoica.
\newblock {SpecInfer}: Accelerating generative {LLM} serving with tree-based
  speculative inference and verification.
\newblock In \emph{ASPLOS}, 2024.

\bibitem{dao2022flashattention}
T.~Dao, D.~Y.~Fu, S.~Ermon, A.~Rudra, and C.~R{\'e}.
\newblock Flash{A}ttention: Fast and memory-efficient exact attention with
  {IO}-awareness.
\newblock In \emph{NeurIPS}, 2022.

\bibitem{dao2023flashattention2}
T.~Dao.
\newblock Flash{A}ttention-2: Faster attention with better parallelism and work
  partitioning.
\newblock In \emph{ICLR}, 2024.

\bibitem{ye2025flashinfer}
Z.~Ye, L.~Zheng, Q.~Sun, W.~Wu, J.~Zhang, Z.~Chen, and I.~Stoica.
\newblock {FlashInfer}: Efficient and customizable attention engine for {LLM}
  inference serving.
\newblock \emph{arXiv:2501.01005}, 2025.

\bibitem{yao2024deft}
Y.~Yao, S.~Zuo, S.~Shi, X.~Ning, K.~Wang, Z.~Lu, Y.~Li, and Y.~Wang.
\newblock {DeFT}: Decoding with flash tree-attention for efficient
  tree-structured {LLM} inference.
\newblock \emph{arXiv:2404.00242}, 2024.

\bibitem{tillet2019triton}
P.~Tillet, H.-T.~Kung, and D.~Cox.
\newblock Triton: An intermediate language and compiler for tiled neural
  network computations.
\newblock In \emph{MAPL}, 2019.

\bibitem{zaheer2020bigbird}
M.~Zaheer, G.~Guruganesh, K.~A.~Dubey, J.~Ainslie, C.~Alberti, S.~Ontanon,
  P.~Pham, A.~Ravula, Q.~Wang, L.~Yang, and A.~Ahmed.
\newblock {B}ig{B}ird: Transformers for longer sequences.
\newblock In \emph{NeurIPS}, 2020.

\bibitem{beltagy2020longformer}
I.~Beltagy, M.~E.~Peters, and A.~Cohan.
\newblock Longformer: The long-document transformer.
\newblock \emph{arXiv:2004.05150}, 2020.

\bibitem{kwon2023efficient}
W.~Kwon, Z.~Li, S.~Zhuang, Y.~Sheng, L.~Zheng, C.~H.~Yu, J.~Gonzalez,
  H.~Zhang, and I.~Stoica.
\newblock Efficient memory management for large language model serving with
  {PagedAttention}.
\newblock In \emph{SOSP}, 2023.

\bibitem{sharegpt}
Anon8231489123.
\newblock {ShareGPT\_Vicuna\_unfiltered}: A dataset of human--{GPT}
  conversations.
\newblock \url{https://huggingface.co/datasets/anon8231489123/ShareGPT_Vicuna_unfiltered},
  2023.

\end{thebibliography}

\end{document}